What We Eval in the Shadows

### 2021-09-09 paper/main.tex
6

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\renewcommand{\k}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother
\lstdefinelanguage{smalleR} {
  keywords={
    for,
    if,
    else,
    function
  },
  sensitive=true, % keywords are case-sensitive
  morecomment=[l]{\#}, % l is for line comment
  morestring=[b]{"} % defines that strings are enclosed in double quotes
}
\lstset{
  language={smalleR},
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
%  framexleftmargin=1mm,
%  framexrightmargin=1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  showspaces=false,
  showstringspaces=false,
  deletekeywords={see_if, $, _, .},
  % alsoletter={., _, $},
  %emph={eval.parent, evalq, eval, local},
  %emphstyle=\underbar,
  breaklines=true
}

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute}, and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected the top 117 sites, which contribute to 90\% of the calls in
the corpus. We present examples from nine classes from this batch. The
classification is not exhaustive; we present results from only 117 of the total
\PkgEvalCallSites call sites in our corpus, and many uses of \eval do not fit in
a specific category. The \eval usage patterns are described below.

\paragraph{Behavioral Extension} Sometimes it is convenient to allow users to
extend the behavior of a library function. \Eval allows clients to provide a
simple text string that is executed in the context of the function. The
example here allows users to provide formulas as strings, such as
\k{"exp(-x^2/2)"}, the contract being that the text can refer to some variable
\k{x} that will be set up by the called function.\footnote{The function is used
to generate a sequence of random numbers using the adaptive rejection sampling
algorithm and the string formula is used to describe the target density.}

% To remove "Listing i" in the captions from this point, as we do not refer to them by number
% We define and refer to listings in 2.2 so we need to put that captionsetup only here
\captionsetup[lstlisting]{labelformat=empty} 
\begin{lstlisting}[caption={\k{AdapSamp::rARS}}, captionpos=b]
 function(n,formula) {
   p <- function(x) eval(parse(text=formula))
\end{lstlisting}

\paragraph{Code Generation}
\Eval can be used to generate boilerplate code that would either be tedious to
write by hand or that is only known after deployment. The following code snippet
for the \k{Rcpp} package uses \eval to automatically generate R bindings to C++
methods.

\begin{lstlisting}[caption={\k{Rcpp::.makeCppMethods}}, captionpos=b]
 function(methods,env) {
   for(what in cppMethods)
     methods[[what]] <-
       eval(substitute(function(...)CppObject$what(...)), env)
\end{lstlisting}

\paragraph{Extracting Varargs}
\Eval is used to obtain the source code of the arguments passed to a vararg
(\k{...}) parameter. In the following example, \k{substitute(alist(...))} yields
an expression representing a call to \k{alist} with the \k{...} expanded to the
arguments supplied to the function. \Eval evaluates this expression and returns
a \k{list} of the unevaluated argument expressions.

\begin{lstlisting}[caption={\k{statnet.common::NVM}}, captionpos=b]
 function(...) {
   exprs <- eval(substitute(alist(...)))
\end{lstlisting}\medskip

\paragraph{Controlled Evaluation}
\Eval can be used to control when and if expressions are evaluated. For example,
\k{assertthat} provides users with the ability to make assertions and, if they
fail, produce legible error messages. The \k{assertthat::see_if} function
accepts a vararg with a list of expressions. It extracts these expressions
(\emph{Extracting Varargs} pattern) and evaluates them sequentially in the
caller's environment using \k{eval}. If an expression does not hold, an error is
generated.

\begin{lstlisting}[caption={\k{assertthat::see\_if}}, captionpos=b]
 function(..., env=parent.frame(), msg=NULL) {
   asserts <- eval(substitute(alist(...)))
   for (assertion in asserts) {
     res <- tryCatch({
       eval(assertion, env)
     }, assertError=function(e) { structure(FALSE, msg=e$message) })
\end{lstlisting}\medskip

\begin{lstlisting}[caption={\k{igraph::do\_call}}, captionpos=b]
 function(f,...,args=list(),env=parent.frame()) {
   f <- substitute(f)
   eval(make_call(f,...,args), env)
 }
\end{lstlisting}\medskip

\begin{lstlisting}[caption={\k{survival::survfit.formula}}, captionpos=b]
 function(formula, data, weights, ...) {
   Call <- match.call()
   indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
   temp <- Call[c(1, indx)]
   temp[[1L]] <- quote(stats::model.frame)
   mf <- eval.parent(temp)
\end{lstlisting}\medskip

\begin{lstlisting}[caption={\k{base::write.csv}}, captionpos=b]
 function (...) {
   Call <- match.call(expand.dots = TRUE)
   Call$sep <- ","
   Call$dec <- "."
   Call$qmethod <- "double"
   Call[[1L]] <- as.name("write.table")
   eval.parent(Call)
 }
\end{lstlisting}\medskip

\paragraph{Non-standard Scoping}

\Eval allows functions to resolve symbols in different environments than the
current one. For example, the \k{data.table} package redefines the subsetting
operator, \k{[}\footnote{Defined as a triplet \texttt{[i, j, k]}, where
  expression \texttt{j} is used to calculate a subset of rows from \texttt{i}
  grouped by \texttt{k}.}, to enable compact queries on data frames. The example
below shows the use of this operator on the \k{flights} data frame to compute
the average departure delays of Delta flights in 2014 by the origin airport. The
symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up in
the current environment, but in the \k{flights} data frame.

@@

Non-standard scoping is an important use case of \eval in R. It enables the
design of functions that require less typing. Since R is used primarily
interactively, this trade-off is justifiable.

@@

The \k{glue} package contains a variant of \k{glue} function that leverages
non-standard scoping for interpolation in a custom environment. For example, the
following snippet pipes the subsetted data frame into the \k{glue_data} function
that uses it to resolve the symbols for string interpolation.

\begin{lstlisting}[caption={\k{glue::glue_data}}, captionpos=b]
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |>
    glue_data("{origin} has average delay of {round(V1)} min")
# LGA has average delay of 11 min
# EWR has average delay of 15 min
# JFK has average delay of 12 min
\end{lstlisting}\medskip

@@

\begin{lstlisting}[caption={\k{aibd::scalaEnsure}}, captionpos=b]
 function() {
   env <- parent.env(environment())
   eval(parse(text=paste0('unlockBinding("s", env)')))
   assign("s", s, envir=env)
   lockBinding("s", env)
\end{lstlisting}\medskip

@@

The \eval function is used widely, and in varied ways, in R. The function is an
essential tool for language implementers. Our analysis observed that the base
libraries heavily depend on \eval, and any R program will end up calling it
through the core of the language. Independently developed libraries often use
\eval in subtle ways, mostly to perform macro-programming tasks. Our review of
code written by less experienced users suggests that \eval is exceedingly rarely
used. So, it is fair to conclude that a majority of R programmers never have to
write a call to \eval, but that the code they write would not run without it.

>>>>

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\renewcommand{\k}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother
\lstdefinelanguage{smalleR} {
  keywords={
    for,
    if,
    else,
    function,
    break,
    in
  },
  sensitive=true, % keywords are case-sensitive
  morecomment=[l]{\#}, % l is for line comment
  morestring=[b]{"} % defines that strings are enclosed in double quotes
}
\lstset{
  language={smalleR},
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
%  framexleftmargin=1mm,
%  framexrightmargin=1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  showspaces=false,
  showstringspaces=false,
  deletekeywords={see_if, $, _, .},
  % alsoletter={., _, $},
  %emph={eval.parent, evalq, eval, local},
  %emphstyle=\underbar,
  breaklines=true
}

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute}, and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected the top 117 sites, which contribute to 90\% of the calls in
the corpus. We present examples from nine classes from this batch. The
classification is not exhaustive as we only study a fraction from the
\PkgEvalCallSites call sites in our corpus, and many uses of \eval do not fit
in a specific category. Instead, the goal is to demonstrate the wide variety of
uses to which \eval is put to use. At the high level, the usage can be grouped
into two groups: one contains patterns that aim to provide better
API for R users and the other focuses on simplifying implementation.

\subsection{Better API}

This category contains \eval used to develop a better API in terms of
ergonomics, extensibility, and ease of use. This is important given
that R is often being used interactively for data exploration.

\paragraph{Non-standard Scoping} \Eval allows functions to resolve symbols in
different environments than the current one. For example, the \k{data.table}
package redefines the subsetting operator, \k{[}\footnote{Defined as a triplet
  \texttt{[i, j, k]}, where expression \texttt{j} is used to calculate a subset
of rows from \texttt{i} grouped by \texttt{k}.}, to enable compact queries on
data frames. The example below shows the use of this operator on the
\k{flights} data frame to compute the average departure delays of Delta flights
in 2014 by the origin airport. The symbols \k{carrier}, \k{year},
\k{dep_delay}, and \k{origin} are not looked up in the current environment but
in the \k{flights} data frame.

@@

Non-standard scoping is an important use case of \eval in R. It enables the
design of functions that require less typing and syntactic noise at the cost of
some ambiguity. Since R is used primarily interactively, this trade-off is
justifiable.

@@

The \k{glue} package contains a variant of \k{glue} function that leverages
non-standard scoping for interpolation in a custom environment. For example,
the following snippet pipes the subsetted data frame into the \k{glue_data}
function used to resolve the symbols for string interpolation.

\begin{lstlisting}[caption={\k{glue::glue_data}}, captionpos=b]
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |>
    glue_data("{origin} has average delay of {round(V1)} min")
# LGA has average delay of 11 min
# EWR has average delay of 15 min
# JFK has average delay of 12 min
\end{lstlisting}

\paragraph{Behavioral Extension} Sometimes it is convenient to allow users to
extend the behavior of a library function. \Eval allows clients to provide a
simple text string that is executed in the context of the function. The example
here allows users to provide formulas as strings, such as \k{"exp(-x^2/2)"},
the contract being that the text can refer to some variable \k{x} the called
function will set up.\footnote{The function is used to generate a sequence of
  random numbers using the adaptive rejection sampling algorithm and the string
formula is used to describe the target density.}

% To remove "Listing i" in the captions from this point, as we do not refer to them by number
% We define and refer to listings in 2.2 so we need to put that captionsetup only here
\captionsetup[lstlisting]{labelformat=empty} 
\begin{lstlisting}[caption={\k{AdapSamp::rARS}}, captionpos=b]
 function(n,formula) {
   p <- function(x) eval(parse(text=formula))
   ...
 }
\end{lstlisting}

\paragraph{Extracting Varargs} \Eval is used to obtain the source code of the
arguments passed to a vararg parameter (\k{...}). Concretely,
\k{substitute(alist(...))} yields an expression representing a call to
\k{alist} with the \k{...} expanded to the arguments supplied to the function.
Passing this to \eval returns a list of the unevaluated argument
expressions. In the following example, it is used to find and return the value of the first expression that evaluates to non-null value. The extracting varargs pattern is needed to avoid premature promise forcing.

\begin{lstlisting}[caption={\k{statnet.common::NVL}}, captionpos=b]
 function(...) {
   for (e in eval(substitute(alist(...)))) {
     if (!is.null(eval(e, parent.frame()))) break
   }
   e
 }
\end{lstlisting}

\paragraph{Controlled Evaluation} \Eval can be used to control when and if
expressions are evaluated. For example, \k{assertthat} provides users with the
ability to make assertions and, if they fail, produce legible error messages.
The \k{assertthat::see_if} function accepts a vararg with a list of
expressions. It extracts these expressions using the extracting varargs
pattern and evaluates them sequentially in the caller's environment using
\k{eval}. If an expression does not hold, an error is generated.

\begin{lstlisting}[caption={\k{assertthat::see_if}}, alsoletter={\_}, captionpos=b]
 function(..., env=parent.frame(), msg=NULL) {
   asserts <- eval(substitute(alist(...)))
   for (assertion in asserts) {
     res <- tryCatch({
       eval(assertion, env)
     }, assertError=function(e) { structure(FALSE, msg=e$message) })
   ...
 }
\end{lstlisting}

\begin{lstlisting}[caption={\k{igraph::do\_call}}, captionpos=b]
 function(f,...,args=list(),env=parent.frame()) {
   f <- substitute(f)
   eval(make_call(f,...,args), env)
 }
\end{lstlisting}

\subsection{Implementation simplification}

The patterns in this category use \eval to simplify implementation in terms of
avoiding repetitive and boilerplate code.

\paragraph{Code Generation} \Eval can be used to generate code that would
either be tedious to write by hand or that is only known after deployment. The
following code snippet for the \k{Rcpp} package uses \eval to automatically
generate R bindings to C++ methods.

\begin{lstlisting}[caption={\k{Rcpp::.makeCppMethods}}, captionpos=b]
 function(methods, env) {
   for(what in cppMethods)
     methods[[what]] <-
       eval(substitute(function(...)CppObject$what(...)), env)
   ...
 }
\end{lstlisting}

\begin{lstlisting}[caption={\k{survival::survfit.formula}}, captionpos=b]
 function(formula, data, weights, ...) {
   Call <- match.call()
   indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
   temp <- Call[c(1, indx)]
   temp[[1L]] <- quote(stats::model.frame)
   mf <- eval.parent(temp)
   ...
 }
\end{lstlisting}
\medskip

\begin{lstlisting}[caption={\k{base::write.csv}}, captionpos=b]
 function (...) {
   Call <- match.call(expand.dots = TRUE)
   Call$sep <- ","
   Call$dec <- "."
   Call$qmethod <- "double"
   Call[[1L]] <- as.name("write.table")
   eval.parent(Call)
 }
\end{lstlisting}

@@

\begin{lstlisting}[caption={\k{aibd::scalaEnsure}}, captionpos=b]
 function() {
   env <- parent.env(environment())
   eval(parse(text=paste0('unlockBinding("s", env)')))
   assign("s", s, envir=env)
   lockBinding("s", env)
   ...
 }
\end{lstlisting}\medskip

@@

The \eval function is used widely, and in varied ways in R. The function is an
essential tool for language implementers. Our analysis observed that the base
libraries heavily depend on \eval, and any R program will end up calling it
through the core of the language. Independently developed libraries often use
\eval in subtle ways, mostly to perform macro-programming tasks. Our review of
code written by less experienced users suggests that \eval is exceedingly rarely
used. So, it is fair to conclude that most R programmers never have to
write a call to \eval, but the code they write would not run without it.
